### 2025-05-21 docs/article/Sections/state-of-the-art.tex
chapter 2 - added progressive rendering

>>>>

Pursuing the optimal user experience for web users has been a consistent goal
since the inception of the World Wide Web.
Consequently, various approaches have been explored to deliver the initial
meaningful content to the end-user as promptly as possible, while the remaining
content \textit{progressively} (or incrementally) loads as the server streams
the HTML content.

\textit{\textbf{Progressive rendering}} and \textit{\textbf{progressive loading}}
encompass different concepts. The former pertains to the \textit{dynamic} content of a
\textit{dynamic web page}, encompassing elements with logic and placeholders
that are fulfilled by data from an object \textit{model} constructed at runtime. 
On the other hand, the latter is associated with \textit{render-blocking
resources} such as scripts, stylesheets, and HTML imports, which may hinder the
browser from rendering page content to the screen~\cite{progressive-2022}.
The notion of \textit{progressive rendering} has typically been aligned with
\textit{client-side rendering} (CSR), where a single HTML page with static
content is delivered upfront, while the dynamic content is fetched as the data
becomes available to complete the web page.
However, despite being overlooked, HTTP and browsers were designed from their
inception to also support this feature in the context of SSR approaches,
which offers the advantage of not being dependent on
\textit{rendering-blocking resources}, such as the required JavaScript for CSR.
% \textit{Progressive server-side rendering} (PSSR) is the
% ability of streaming HTML content to the client in chunks or data frames, as it
% gets resolved on the server at runtime. 
% This lets the user-interface to be rendered incrementally (i.e.
% \textit{progressively}) by the browser in line with the availability of the data. 
% However, most SSR template engines do not inherently support this feature at the
% template level, requiring developers to manually break down web pages into 
% fragments or partials to enable PSSR. 
An example of this limitation in SSR web templates
was highlighted by Jeff Atwood in 2005, who criticized Microsoft ASP.NET for loading the
entire web page into memory before sending any data to the browser~\cite{pssr2005}.
% This practice delays the display of content to the user until the server
% completes its processing, contradicting the original design intent of HTML to
% render progressively as content is received. 
% Notably, both Netscape and Internet Explorer were capable of rendering partial
% HTML content from their inception.
Despite historical critiques and HTML's inherent capabilities, most Web application
frameworks, such as ASP.Net, Express.js, Spring, and others, 
persistently lack support for progressive rendering, leading to the
appearance of alternative techniques leveraging client-side JavaScript.
Since 2007, various patents have addressed the PSSR issue, with Microsoft's
patent enabling the infinite scrolling technique by displaying a single page of
results~\cite{scroll2007}, and Yahoo's patent focusing on differentiating
elements based on their position relative to the visible
portion~\cite{schiller2007progressive}.

Former techniques all describe methods of progressively adding content to a web
page depending on client-side JavaScript.

Introduce Thymeleaf

Spring resolver does ot natively

Carvalho~\cite{carvalho2023async} introduced an SSR solution that manages
multiple data models and asynchronous APIs while ensuring well-formed HTML,
PSSR, and non-blocking template resolution.
That solution builds upon HtmlFlow, a Java DSL for HTML, with an internal
processing mechanism that streams HTML in chunks as data items become available
from data sources, without requiring client-side JavaScript.
Like Thymeleaf, their proposal avoids the use of client-side JavaScript;
however, unlike Thymeleaf, it supports a wide range of asynchronous APIs and
multiple data sources, not limited to the \texttt{Publisher} API.
However, the main counter-argument is the non-trivial management of the
\emph{resume} callback in continuations~\cite{von2003events,callbackhell}, which
is used to linearize the execution flow between asynchronous calls.
